Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-14 bounds-files/sec03-approx-kde.tex
Remove negative vspace and remove NIPS formatting.
Refactor KDE a bit; we can write about both approximation schemes.

Ram et.~al. present a clever technique for bounding the
running time of approximate kernel density estimation based on the properties of
the kernel, when the kernel is shift-invariant and satisfies a few assumptions
\cite{ram2009}.  We will restate these assumptions and provide an adapted proof
using Theorem \ref{thm:ct-runtime}, which gives a tighter bound.

Approximate kernel density estimation is a common application of dual-tree
algorithms \cite{gray2003nonparametric, nbody}.  Given a query set $S_q$, a
reference set $S_r$ of size $N$, and a kernel function $\mathcal{K}(\cdot,
\cdot)$, the goal is to calculate, for each $p_q \in S_q$, a density estimate
$f(p_q)$ such that $| f(p_q) - f^*(p_q) | < N \epsilon$ for some
parameter $\epsilon$ and

%\vspace*{-0.5em}
\begin{equation}
f^*(p_q) = \sum_{p_r \in S_r} \mathcal{K}(p_q, p_r).
\end{equation}
%\vspace*{-0.5em}

This is related to the well-studied problem of kernel summation, which can
also be solved with dual-tree algorithms \cite{lee2006faster, lee2008fast}.  In
both of those problems, simple geometric observations can be made to accelerate
computation: when $\mathcal{K}(\cdot, \cdot)$ is shift-invariant, faraway points
have very small kernel evaluations.  Thus, trees can be built on $S_q$ and
$S_r$, and node combinations can be pruned when the nodes are far apart while
still obeying the error bound $N \epsilon$.\footnote{Here we consider
only absolute-value approximation.  Our methods can be easily adapted to other
schemes, such as relative-value approximation.  \cite{ram2009} considers
relative-value approximation in far more detail.}

@@

\vspace*{-0.8em}
\begin{equation}
\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))
\ge \epsilon.
\end{equation}

@@

\vspace*{-1.5em}
\begin{equation}
d_{\max}(\mathscr{N}_q, \mathscr{N}_r) - d_{\min}(\mathscr{N}_q, \mathscr{N}_r)
= d(p_q, p_r) + 2^{s_r^{\max} + 1} - d(p_q, p_r) + 2^{s_r^{\max} + 1} = 2^{s_r^{\max} + 2}.
\end{equation}
\vspace*{-0.8em}

@@

%\vspace*{-1.0em}
%\begin{eqnarray}
%B_{S_r}\left(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}\right) &\subseteq&
%B_{S_r}\left(p_r, 2 (\mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1})\right) \\
% &=& B_{S_r}\left(p_r, 2^{s_r^{\max} + 2} \left(
%\frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1}} + 1\right)\right).
%\label{eqn:kde_ball}
%\end{eqnarray}
%\vspace*{-0.5em}

@@

%\vspace*{-0.8em}
%\begin{equation}
%\zeta = (1/4) ( \epsilon^{-1} \ln(\epsilon^{-1} - 1) +
%(\beta \alpha) / \epsilon ).
%\end{equation}
%\vspace*{-0.8em}
%
%Now, consider the problem of range count, where the number of points in the
%range $[0, u]$ for each query point $p_q \in S_q$ is desired.  If the sigmoidal
%kernel is used with $\beta = u$ for approximate kernel density estimation, then
%as $\alpha \to 0$ and $\epsilon \to 0$, the results will converge to the result
%for exact range count.  Therefore, when $\alpha > 0$ or $\epsilon > 0$, the
%result is approximate range count, and we can use the results of Theorem
%\ref{thm:kde-bound} to show a runtime bound.
%
%Suppose, for the sake of simplicity, that $\alpha = O(\epsilon^{-1})$; then,
%$\zeta = O(\epsilon^{-1} \log \epsilon^{-1} + \epsilon^{-2} u) = O(\epsilon^{-2}
%u)$.  Then, the runtime of approximate range count is bounded as
%
%\vspace*{-1em}
%\begin{eqnarray}
%\begin{equation}
%O(c_r^{8 + \log_2(\epsilon^{-2} u)} \nu N) = O(c_r^{8 + 2 \log_2(\epsilon^{-1}) +
%\log_2(u)} \nu N)
% = O(\epsilon^{-1} u c_r^{8 + 3 \log_2 c_r} \nu N).
%\end{equation}
%\end{eqnarray}
%\vspace*{-1.5em}

>>>>

\citet{ram2009} present a clever technique for bounding the
running time of approximate kernel density estimation based on the properties of
the kernel, when the kernel is shift-invariant and satisfies a few assumptions.
We will restate these assumptions and provide an adapted proof using Theorem
\ref{thm:ct-runtime}, which gives a tighter bound.

Approximate kernel density estimation is a common application of dual-tree
algorithms \citep{gray2003nonparametric, nbody}.  Given a query set $S_q$, a
reference set $S_r$ of size $N$, and a kernel function $\mathcal{K}(\cdot,
\cdot)$, the true kernel density estimate for a query point $p_q$ is given as

\begin{equation}
f^*(p_q) = \sum_{p_r \in S_r} \mathcal{K}(p_q, p_r).
\end{equation}

There are two simple ways to approximate $f^*(p_q)$: {\it absolute value
approximation} and {\it relative value approximation}.  Absolute value
approximation requires that each density estimate $f(p_q)$ is within $\epsilon$
of the true estimate $f^*(p_q)$:

\begin{equation}
| f(p_q) - f^*(p_q) | < N \epsilon.
\end{equation}

Relative value approximation is a more flexible approximation scheme; given some
parameter $\epsilon$, the requirement is that each density estimate is within a
relative tolerance of $f^*(p_q):$

\begin{equation}
\frac{| f(p_q) - f^*(p_q) |}{| f^*(p_q) |} < N \epsilon.
\end{equation}

Kernel density estimation is related to the well-studied problem of kernel
summation, which can also be solved with dual-tree algorithms
\citep{lee2006faster, lee2008fast}.  In both of those problems, regardless of
the approximation scheme, simple geometric observations can be made to
accelerate computation: when $\mathcal{K}(\cdot, \cdot)$ is shift-invariant,
faraway points have very small kernel evaluations.  Thus, trees can be built on
$S_q$ and $S_r$, and node combinations can be pruned when the nodes are far
apart while still obeying the error bounds.

In the following two subsections, we will separately consider both the absolute
value approximation scheme and the relative value approximation scheme.

\subsection{Absolute value approximation}

@@

\begin{equation}
\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))
\ge \epsilon.
\end{equation}

@@

\begin{equation}
d_{\max}(\mathscr{N}_q, \mathscr{N}_r) - d_{\min}(\mathscr{N}_q, \mathscr{N}_r)
= d(p_q, p_r) + 2^{s_r^{\max} + 1} - d(p_q, p_r) + 2^{s_r^{\max} + 1} = 2^{s_r^{\max} + 2}.
\end{equation}

@@

%\begin{eqnarray}
%B_{S_r}\left(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}\right) &\subseteq&
%B_{S_r}\left(p_r, 2 (\mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1})\right) \\
% &=& B_{S_r}\left(p_r, 2^{s_r^{\max} + 2} \left(
%\frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1}} + 1\right)\right).
%\label{eqn:kde_ball}
%\end{eqnarray}

@@

%\begin{equation}
%\zeta = (1/4) ( \epsilon^{-1} \ln(\epsilon^{-1} - 1) +
%(\beta \alpha) / \epsilon ).
%\end{equation}
%
%Now, consider the problem of range count, where the number of points in the
%range $[0, u]$ for each query point $p_q \in S_q$ is desired.  If the sigmoidal
%kernel is used with $\beta = u$ for approximate kernel density estimation, then
%as $\alpha \to 0$ and $\epsilon \to 0$, the results will converge to the result
%for exact range count.  Therefore, when $\alpha > 0$ or $\epsilon > 0$, the
%result is approximate range count, and we can use the results of Theorem
%\ref{thm:kde-bound} to show a runtime bound.
%
%Suppose, for the sake of simplicity, that $\alpha = O(\epsilon^{-1})$; then,
%$\zeta = O(\epsilon^{-1} \log \epsilon^{-1} + \epsilon^{-2} u) = O(\epsilon^{-2}
%u)$.  Then, the runtime of approximate range count is bounded as
%
%\begin{eqnarray}
%\begin{equation}
%O(c_r^{8 + \log_2(\epsilon^{-2} u)} \nu N) = O(c_r^{8 + 2 \log_2(\epsilon^{-1}) +
%\log_2(u)} \nu N)
% = O(\epsilon^{-1} u c_r^{8 + 3 \log_2 c_r} \nu N).
%\end{equation}
%\end{eqnarray}
